Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination

### 2024-01-14 manuscript/manuscript.Rmd
Reorganized folder

Seeking and giving advice can be challenging conversations to navigate.
Advice seekers may be embarrassed to ask for help [@Benabou2022], and
they may fail to do so if the information they receive could lead to
disappointment [@Gill2012]. Indeed, people prefer advisers who withhold
unpleasant information [@Shalvi2019] and punish those who tell them
things they do not want to hear [@John2019]. People appear to engage in
strategic selection of advice to avoid unpleasant information and
protect the utility they get from positive beliefs about themselves and
the future they anticipate [@Loewenstein2018; @Loewenstein1987]. But do
advice givers anticipate and consider the hedonic cost of honest
information to the recipient? While previous work has emphasized the
instrumental value of advice (e.g., satisfaction with a decision or
accuracy of a prediction), we focus on the motivations of the person
providing advice.

Research on advice has largely focused on settings that do not tackle
belief utility. For example, participants may have been asked to
estimate the number of balls in an urn or choose from pairs of lotteries
[@Benjamin2015]. Advisers have been shown to be motivated to give
accurate advice [@Jonas2003], and engage in perspective-taking to meet
the preferences of the advisee [@Kray2000; @Kray1999]. When they are
uncertain, they prefer that advisees do not rely (extensively) on their
recommendations [@Ache2020]. Thus, it seems that people are concerned
about providing good advice even when they are not incentivized for the
recipient’s decision quality.

In many real settings, however, advice incorporates beliefs about the
recipient. Someone recommending an expensive restaurant conveys that
they think the advisee is wealthy enough to afford an expensive meal,
and someone recommending a student pursue a PhD in economics assumes
that they are intellectually capable. Conversely, telling a poorly
performing student to switch to a less-demanding major conveys a belief
that the student is not capable of improving. Thus, the painful truth
may prevent someone from making a bad decision, it may come at a cost to
the recipient and damage the relationship between the adviser and
advisee. We propose that advisers take these costs into account, even
when the interactions are entirely anonymous and they are incentivized
for the quality of the advisee’s decision.

@@

Our account therefore predicts that people receive advice that is more
flattering and suggestive of inflated ability and optimistic outcomes
than would advice in the absence of such interpersonal considerations.
However, we propose that this is not true for all advisees equally.
Presenting flattering advice requires the formation of beliefs about the
advisee's expectations (reference point). @Exley2022 show that men are
more overconfident than women, conditional on equal performance and that
people anticipate but fail to incorporate this difference into their
decisions. An adviser who seeks to avoid disappointment and incorporates
expectations into their advice would then present more favorable advice
to male advisees than to female advisees (see Figure 1 for an
illustration). This provides a novel account for a well-known finding
that men receive more aspirational advice than do women [@Kanze2018].

Notably, when it pays to have accurate beliefs, such flattering advice
would come at a cost: while men would be more likely to aim higher, they
would also suffer greater costs from aiming too high and failing to meet
the objective. This may seem at odds with a gender gap favoring men.
However, observational data often miss people who were unsuccessful. For
example, job candidates who ask for a high starting salary may be less
likely to receive a job offer, but will also have a higher starting
salary conditional on receiving an offer.

In a large, preregistered experiment (n = 2,000), we show that advisers
consider the expectations of advisees, even when such expectations do
not provide informative insights for the underlying decision-making
problem. We find that as a result, showing expectations (but not gender)
to advisers leads men to receive more aspirational advice than women.
Incorporating expectations leads to worse advice, even when advisers are
incentivized based on the outcome of the advisee’s decision. In a second
preregistered experiment (n = 700), we show that advisers respond to
interpersonal considerations. They advice people that they are more
attractive than they believe them to be, and inflate to a greater extent
when incentivized for likability rather than accuracy. Advisees indeed
reward such flattering advice, finding the advisers more likeable and no
less trustworthy than those who provide more accurate advice.

@@

We report all sample sizes, data exclusions, all manipulations, and all
measures in the studies. Screen captures of the experimental materials
are available in the Supplemental Information. The complete data, code
to reproduce all statistical analyses and figures in the manuscript, as
well as the preregistration reports are available via OSF. All our
studies were preregistered on AsPredicted.org. (will do this later)

>>>>

Seeking and giving advice can present conversational challenges that
are difficult to navigate. Advice seekers may be embarrassed to ask for
help [@Benabou2022], and they may fail to do so if the information they
receive could lead to disappointment [@Gill2012]. Indeed, people prefer
advisers who withhold unpleasant information [@Shalvi2019] and punish
those who tell them things they do not want to hear [@John2019]. People
appear to engage in strategic selection of advice to avoid unpleasant
information and protect the utility they get from positive beliefs about
themselves and the future they anticipate [@Loewenstein2018;
@Loewenstein1987]. But do advice givers anticipate and consider the
hedonic cost of honest information to the recipient? While previous work
has emphasized the instrumental value of advice (e.g., satisfaction with
a decision or accuracy of a prediction), we focus on the motivations of
the person providing advice.

Research on advice has largely focused on settings that do not tackle
belief utility. For example, participants may have been asked to
estimate the number of balls in an urn or choose from pairs of lotteries
[@Benjamin2015]. Advisers have been shown to be motivated to give
accurate advice [@Jonas2003], and engage in perspective-taking to meet
the preferences of the advisee [@Kray2000; @Kray1999]. When they are
uncertain, they even prefer that advisees do not rely (extensively) on their
recommendations [@Ache2020]. Thus, it seems that people are concerned
about providing good advice even when they are not incentivized for the
recipient’s decision quality.

In many real settings, however, advice incorporates beliefs about the
recipient. Someone recommending an expensive restaurant conveys that
they think the advisee is wealthy enough to afford an expensive meal,
and someone recommending a student pursue a PhD in economics assumes
that they are intellectually capable. Conversely, telling a poorly
performing student to switch to a less-demanding major conveys a belief
that the student is not capable of improving. Thus, while the painful truth
may prevent someone from making a bad decision, it may come at a hedonic cost to
the recipient and damage the relationship between the adviser and
advisee. We propose that advisers take these costs into account, even
when the interactions are entirely anonymous and they are incentivized
for the quality of the advisee’s decision.

@@

Our account therefore predicts that people give and receive advice that is more
flattering and suggestive of inflated ability and optimistic outcomes
than would advice in the absence of such interpersonal considerations.
However, we propose that this is not true for all advisees equally.
Presenting flattering advice requires the formation of beliefs about the
advisee's expectations (reference point). @Exley2022 show that men are
more overconfident than women, conditional on equal performance, and that
people anticipate this difference. An adviser who seeks to avoid disappointment and incorporates
expectations into their advice would then present more favorable advice
to male advisees than to female advisees (see Figure 1 for an
illustration). This provides a novel account for a well-known finding
that men receive more aspirational advice than do women [@Kanze2018].
Notably, when it pays to have accurate beliefs, such flattering advice
would come at a cost: while men would be more likely to aim higher, they
would also suffer greater costs from aiming too high and failing to meet
the objective.

In a large, preregistered experiment (n = 2,000), we show that advisers
consider the expectations of advisees, even when such expectations do
not provide informative insights for the underlying decision-making
problem. We find that as a result, showing expectations (but not gender)
to advisers leads men to receive more aspirational advice than women.
Incorporating expectations leads to worse advice, even when advisers are
incentivized based on the outcome of the advisee’s decision. In a second
preregistered experiment (n = 700), we show that advisers respond to incentives for
interpersonal considerations. They tell people that they are more
attractive than they believe them to be, and inflate to a greater extent
when incentivized for likability rather than accuracy. Notably, they inflate
even when only incentivized for accuracy. Advisees indeed
reward such flattering advice, rating the advisers more likeable and no
less trustworthy than those who provide more accurate advice.

@@

We report all sample sizes, data exclusions, manipulations, and
measures in the studies. Screen captures of the experimental materials
are available in the Supplemental Information. The complete data, code
to reproduce all statistical analyses and figures in the manuscript, as
well as the preregistration reports are available via OSF. All our
studies were preregistered on AsPredicted.org.^[<https://osf.io/8r3d4/?view_only=5ad7bafcd16b4d4ba08bb28b0e2bd02d>]
